Graphical Tests for Power Comparison of Competing Designs

### 2012-03-31 experiment.tex
conclusions, mainly
included the pairwise comparison

% !TEX root = comparison.tex
\section{Results and Discussion}~\label{results}
%
\subsection{Wind Direction and Airport Efficiency}
Figures \ref{fig:time} and \ref{fig:conf} give an overview of the relationship between the three response variables: accuracy, speed, and confidence level.

@@

@@

For evaluating the results, we are using a generalized linear mixed effects modeling  approach \cite{pinheiro:2000} for each one of the three response values, using the {\t R} package {\tt lme4} \cite{bates:2011}. 
For power predictions (cf. table \ref{tbl:correct}) a logistic regression was fitted for the competing designs, including covariates {\it sample size} (2, 4,6, 8, 10, and 24 percent of the data) and {\it shift} in wind direction (offset of 0, 90, 180, and 270 degrees), both as main effects and in two-way interaction effects with design to assess their effect on the power of designs. To adjust for individuals' ability, a random intercept was included in the model.

@@

\begin{tabular}{cl}
\phantom{\includegraphics[width=0.03\linewidth]{Polar_Line.pdf}} & \vspace{-0.035in} \multirow{10}{*}{\hspace{-0.25in}\includegraphics[width=0.9\linewidth]{turk4-designs.pdf}} \\
\includegraphics[width=0.05\linewidth]{Polar_Line.pdf} \\
\includegraphics[width=0.05\linewidth]{Polar_NoLine.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
\phantom{\includegraphics[width=0.015\linewidth]{Polar_Line.pdf}}\\
\includegraphics[width=0.05\linewidth]{Polar_Line.pdf} \\
\includegraphics[width=0.05\linewidth]{Polar_NoLine.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
  \end{tabular} 
  \vspace{0.1in}
   \caption{Power results for two competing designs: polar versus a Euclidean, with and without a reference line. Error bars indicate 95\% Bonferroni adjusted confidence intervals. Plots are facetted by percentage of data used. }
   \label{fig:treatment}
\end{figure}

Overall, the results show huge differences in the power of designs between polar charts and Euclidean: Euclidean designs are significantly more powerful than polar charts, particularly so with small sample sizes. 
 The reference line has surprisingly little influence, but it helps more for  polar charts than for Euclidean charts,
An increase in sample size has a positive impact on power (see figure \ref{fig:treatment} for an overview). Polar charts need a much bigger sample size to see an increase in power -- only at about 24\% of the original data do we see about the same power as for Euclidean charts of a sample size of 2\%.
The  changes in offset are significant -- interestingly, borderline behavior (90 and 270 degrees) does not show a difference between polar and Euclidean charts, whereas an inversion of the wave pattern (first up, then down), does show a difference. The power of Euclidean lineups suffers significantly  from this inversion whereas power of polar charts is unaffected. This is an unexpected finding, but is consistent throughout different lineups in the data. Figure \ref{fig:power} summarizes the results from the model: power predictions ($y$ axis) are shown by sample size ($x$ axis). The thick lines show average power by design for different shifts in wind direction. The thin lines represent power for individuals. What can be seen is the different impact of the offset by design: while  an offset of 270 degrees (the `mountain' pattern) has the highest power in Euclidean charts, it comes out worst in polar charts.

>>>>

% !TEX root = comparison.tex
\section{Results and Discussion}~\label{results}
%
\subsection{Wind Direction and Airport Efficiency}

Figures \ref{fig:time} and \ref{fig:conf} give an overview of the relationship between the three response variables: accuracy, speed, and confidence level.

@@

Figure \ref{fig:treatment} shows a comparison of power for the four competing designs. 95\% confidence intervals are  Bonferroni adjusted for multiple testing. The bottom two confidence intervals in each panel show a 95\% confidence interval of a direct comparison of the Euclidean design versus the polar design with and without a reference line. With the exception of the 2\% sample none of those confidence intervals include the zero, indicating a significantly higher power for the Euclidean design than for the polar design.

\begin{tabular}{cl}
\phantom{\includegraphics[width=0.03\linewidth]{Polar_Line.pdf}} & \vspace{-0.035in} \multirow{10}{*}{\hspace{-0.5in}\includegraphics[width=0.9\linewidth]{turk4-designs.pdf}} \\
\includegraphics[width=0.045\linewidth]{Polar_Line.pdf} \\
\includegraphics[width=0.045\linewidth]{Polar_NoLine.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
\phantom{\includegraphics[width=0.13\linewidth]{Polar_Line.pdf}}\\
\includegraphics[width=0.045\linewidth]{Polar_Line.pdf} \\
\includegraphics[width=0.045\linewidth]{Polar_NoLine.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_Line.pdf} \\
\includegraphics[width=0.043\linewidth]{Euclidian_NoLine.pdf}\\
\phantom{\includegraphics[width=0.13\linewidth]{Polar_Line.pdf}}\\
  \end{tabular} 
  \vspace{0.1in}
   \caption{Power results for two competing designs: polar versus a Euclidean, with and without a reference line; facetting indicates sample size (as percentage of original data). Error bars indicate single Bonferroni adjusted 95\% confidence intervals. The two confidence intervals at the bottom of each panel shows a power comparison of Euclidean versus polar design according to eqn (\ref{eq:comp}). }
   \label{fig:treatment}
\end{figure}

@@

However, all of these considerations are based on the - rather strong - assumption that all individuals have the same ability to detect the data plot from a lineup. In order to allow for individual differences in visual ability, 
 we use a generalized linear mixed effects modeling  approach \cite{pinheiro:2000} for each of the three response values, using the {\tt R} package {\tt lme4} \cite{bates:2011}. 
For power predictions (cf. table \ref{tbl:correct}) a logistic regression was fitted for the competing designs, including covariates {\it sample size} (2, 4,6, 8, 10, and 24 percent of the data) and {\it shift} in wind direction (offset of 0, 90, 180, and 270 degrees), both as main effects and in two-way interaction effects with design to assess their effect on the power of designs. To adjust for individuals' ability, a random intercept was included in the model.

@@

Overall, the results show huge differences in the power of designs between polar charts and Euclidean: Euclidean designs are significantly more powerful than polar charts, particularly so with small sample sizes. 
 The reference line has surprisingly little influence, but it helps more for  polar charts than for Euclidean charts,
An increase in sample size has a positive impact on power. Polar charts need a much bigger sample size to see an increase in power -- only at about 24\% of the original data do we see about the same power as for Euclidean charts of a sample size of 2\%.
The  changes in offset are significant -- interestingly, borderline behavior (90 and 270 degrees) does not show a difference between polar and Euclidean charts, whereas an inversion of the wave pattern (first up, then down), does show a difference. The power of Euclidean lineups suffers significantly  from this inversion whereas power of polar charts is unaffected. This is an unexpected finding, but is consistent throughout different lineups in the data. Figure \ref{fig:power} summarizes the results from the model: power predictions ($y$ axis) are shown by sample size ($x$ axis). The thick lines show average power by design for different shifts in wind direction. The thin lines represent power for individuals. What can be seen is the different impact of the offset by design: while  an offset of 270 degrees (the `mountain' pattern) has the highest power in Euclidean charts, it comes out worst in polar charts.

### 2012-03-31 intro.tex
included the pairwise comparison

%The power of a statistical test is defined as the probability to reject the null hypothesis. Translated to the visual tests, power is the probability to identify the data plot from the lineup. 
Comparing power of competing designs therefore involves comparing percentages of correct responses $\widehat{\pi}_1$ and $\widehat{\pi}_2$. An $\alpha\cdot$100\% confidence interval for this comparison is given as 
\[
\widehat{\pi}_1 - \widehat{\pi}_2 \pm t_{1-\alpha/2, n-1} \sqrt{\widehat{\pi}_1(1-\widehat{\pi}_1)/n_1 + \widehat{\pi}_2(1-\widehat{\pi}_2)/n_2},
\]
where $n$ is the Welch-Satterthwaite \cite{welch:1947} estimate of the degrees of freedom. Note that we use $\widehat{\pi}_i = (x_i+1)/(n_i+1)$ and $n_i+2$ for a better coverage of the confidence interval \cite{agresti:1998}.

>>>>

%The power of a statistical test is defined as the probability to reject the null hypothesis. Translated to the visual tests, power is the probability to identify the data plot from the lineup. 
Comparing power of competing designs therefore involves comparing percentages of correct responses $\widehat{\pi}_1$ and $\widehat{\pi}_2$. An $\alpha\cdot$100\% confidence interval for this comparison is given as 
\begin{equation}\label{eq:comp}
\widehat{\pi}_1 - \widehat{\pi}_2 \pm t_{1-\alpha/2, n-1} \sqrt{\widehat{\pi}_1(1-\widehat{\pi}_1)/n_1 + \widehat{\pi}_2(1-\widehat{\pi}_2)/n_2},
\end{equation}
where $n$ is the Welch-Satterthwaite \cite{welch:1947} estimate of the degrees of freedom. Note that we use $\widehat{\pi}_i = (x_i+1)/(n_i+1)$ and $n_i+2$ for a better coverage of the confidence interval \cite{agresti:1998}. In the case of more than two competing designs, we have to additionally adjust the confidence intervals for multiple testing, e.g. using a Bonferroni adjustment which uses an adjusted  confidence level $\tilde{\alpha} = \alpha/k$, where $k$ are the number of confidence intervals involved (single adjustment) or the number of comparisons anticipated.
